### 2022-06-22 manuscript.tex
replaced rmd with tex; flattened manuscript and stan folders

---
title: "STRATEGIC SOCIAL LEARNING AND THE POPULATION DYNAMICS OF HUMAN BEHAVIOR: THE GAME OF GO"
short_title: "STRATEGIC SOCIAL LEARNING"
date: March 28, 2014
author: Bret A. Beheim$^1$, Calvin Thigpen$^2$, Richard McElreath$^3$
email: beheim@gmail.com
address: $^1$Department of Anthropology, University of New Mexico. $^2$Institute for Transportation Studies, University of California, Davis. $^3$ Department of Anthropology, University of California, Davis
abstract: "Human culture is widely believed to undergo evolution, via mechanisms rooted in the nature of human cognition. A number of theories predict the kinds of human learning strategies, as well as the population dynamics that result from their action. There is little work, however, that quantitatively examines the evidence for these strategies and resulting cultural evolution within human populations. One of the obstacles is the lack of individual-level data with which to link transmission events to larger cultural dynamics. Here, we address this problem with a rich quantitative database from the East Asian board game known as Go. We draw from a large archive of Go games spanning the last six decades of professional play, and find evidence that the evolutionary dynamics of particular cultural variants are driven by a mix of individual and social learning processes. Particular players vary dramatically in their sensitivity to population knowledge, which also varies by age and nationality. The dynamic patterns of opening Go moves are consistent with an ancient, ongoing arms race within the game itself."
---

```{r, echo = FALSE}

load("./inputs/data_result_list.RData")
load("./inputs/model_result_list.RData")

```

@@

<!-- n -->

In what follows, we present an analysis of such a high-resolution ($n=$ `r dres$n_games`) archive of cultural patterns: records from the East Asian game of Go. Over many decades of professional play, we can see particular strategies and counterstrategies be innovated, copied, decline, and vanish altogether. Our results provide strong evidence that, on the average, these trends are the result of strategic social learning decisions by professional Go players as they adopt and abandon variant openings. This learning model, in aggregate, is the basis for an apparent evolutionary arms race that is largely responsible for the oscillations in Go openings among professional players. In this way, we are able to link individual-level strategic learning to aggregate population dynamics.

# Historical Dynamics in the Game of Go

@@

\begin{figure}[t]
  \begin{center}
    \includegraphics[scale=1.3]{./kifu.pdf}
    \caption{A complete record, with SGF coordinates, for a 1978 game between Japanese professionals Takemiya Masaki and Kudo Norio. Black (Takemiya) begins with a Fourfour at coordinate \textit{pd} (top right), with White (Kudo) responding at \textit{dd} with another Fourfour. White won by resignation after 232 moves. This game is one of the first successful uses of the \textit{dd} response to the Move 1 Fourfour, and \textit{dd} has subsequently become the dominant response among professionals (see Fig. \ref{fig:m2FreqsWins}).}
    \label{fig:kifu}
  \end{center}
\end{figure}

@@

Beyond this basic outline of the game, there are two key rules regulating play. First, stones or groups of stones of one player that become completely surrounded on the grid by stones of the other are removed from the board (this is the only time stones are moved, once placed). Second, to prevent the possibility of the game entering an infinite loop, the *ko* rule restricts play so that the same board state cannot appear twice in one game.

@@

The first few moves of game of Go, which allow the most possibilites, are actually the most predictable. The opening period of the game is often characterized by highly normative, canalized play, where players often follow one of a set of memorized move sequences called *joseki* (or *jungsuk* in Korea), thought to grant neither player a major advantage. Once a particular *joseki* pattern has started, professional players will draw on an encyclopedic knowledge of explored routes in deciding their next move.

Despite the conservatism of opening play in professional games, novel variations appear regularly and, if widely copied, become new *joseki* in the canon. Go historian Peter Shotwell writes of the evolution of *joseki*: "Some are hundreds (or thousands) of years old; others are born, die, and then are reborn again as the style of play changes; still others are yet to be born and patiently wait their turn to be discovered in some flash of ingenuity or desperation." \citep{shotwell2011go}.

<!-- 44's correlation between performance and frequency -->
<!-- time trends in each -->

\begin{figure}[t]
  \begin{center}
    \includegraphics[scale=1.2]{./m1FreqsWins.pdf}
    \caption{The ebb and flow of the three major first move variants, with proportion of games won. The Fourfour's rise and fall, here broken into four eras, closely tracks with its relative performance (\textit{top}) since 1967 ($r=$ `r dres$frq_win_cor_44`). Similarly, the Threethree's brief efflorescence and extinction closely match its relative win rate. Because of its initial dominance, all win rates are expressed relative to the Threefour's, and the dotted gray line indicates the location of the 50\% (break-even) point.}
    \label{fig:m1FreqsWins}
  \end{center}
\end{figure}

<!-- variant frequency closely tracks relative performance -->

\begin{figure}[t]
  \begin{center}
    \includegraphics[scale=1.2]{./m2FreqsWins.pdf}
    \caption{Frequency and performance for four major M2 responses to the Fourfour. Performance for each is calculated as the win rate subtracting the average win rate of alternatives. As at M1, a variants's frequency closely tracks its relative performance. To compare with Fig. \ref{fig:m2FreqsWins}, we define four eras (I to IV) corresponding to the Fourfour's rise and fall.}
    \label{fig:m2FreqsWins}
  \end{center}
\end{figure}

@@

<!-- 1977 had 44 in 78% of games plaed -->
<!-- 1996 had 44 in 82% of games plaed -->

After a brief enthusiasm for the Threethree in the 1960s, the Fourfour rose to dominance, peaking in 1977 (appearing in 78% of games played) and again in 1996 (82% of games played). The primary goal of this analysis is to understand why this pattern exists, and what processes are driving evolution at M1.

We can trace the rapid diffusion of the Fourfour during this time back to the success of a single innovator, Japanese professional Takemiya Masaki. Beginning his pro career at the age of 14 in the 1960s, Takemiya has used the Fourfour almost exclusively; of his 520 games as Black in the database, only six do not start with the Fourfour. Takemiya's novel use of this variant in 1968 marks the beginning of its spread within the broader Go community, and his personal performance for the next five years (64%) also exceeded the average Black win rate (61.5%).

<!-- takemiya has 520 games in the data -->
<!-- takemiya's win rate of 64% vs avg black of 61.5% -->

<!-- before 1970 whites response almost always a threefour on adjacent corner -->
<!-- by 1977 dc response was about 30% of games -->
<!-- db and cq consistently outperfomed dc in the 1970s -->

Subsequent to Takemiya's arrival, the success of the Fourfour appears to have prompted an innovation race in White's Move 2 response. Before 1970, White's response to the Fourfour was almost always a Threefour on an adjacent corner. If we imagine the M1 Fourfour is played in the upper right quadrant (SGF coordinate *pd*), nearly 80% of the time the M2 response in the late 1960s was coordinate *dc* (Fig. \ref{fig:m2FreqsWins}). By 1977, at the Fourfour's first peak, the *dc* response declined to about 30% of games. In the interim, M2 variants at *dp*, *cq* and *dd* all became established in rapid sequence. As measured by the proportion of games White won, *dp* and *cq* both consistently outperformed *dc* in the 1970s, blunting the Fourfour's effectiveness and anticipating its decline in use after 1977.

## Strategic Social Learning

@@

# Methods

@@

# Results

@@

<!-- model intercept of 53.2% -->
<!-- recent personal use predicts current use; 60% use outcome is 58.7% (OR 9.39) ceteris paribus -->
<!-- recent win rate predicts current use, OR of 2.77 -->
<!-- recent personal use and win rate interaction effect exists -->

With all predictors centered on means, the intercept estimates that an average Go player will use the Fourfour about `r mres$pr_44_baseline`% of the time during the study period (i.e. logit(.524)=0.10). A player's recent use of a particular opening move strongly predicts their future use; for an otherwise-average player who used the Fourfour about 60% of their recent games, the probability they will use the Fourfour rises to `r mres$pr_44_b_use_60`% (OR: `r mres$or_44_b_use`). There is a smaller (OR: 2.77), but unambiguous, positive performance effect from past success with the Fourfour, adding about two percentage points to the probability of use for a player who has won 10% more games using the Fourfour than his overall win rate. These two personal effects also interact, so that past success enhances the predictive power of past use (or vis versa).

<!-- recent pop use of 60% means expected prob is 59%, OR 10.475 -->
<!-- recent pop use of 60% and +10% performance, its 61.4% -->

Most important to the question of social learning and cultural evolution, though, is whether there is predictive value in the recent behavior of *other* players, controlling for individual behavior. Indeed, when the Fourfour appears in 60% of *all* recent games, the expected probability of use by an average player rises from `r mres$pr_44_baseline`% to `r mres$pr_44_pop_use_60`% (OR: `r mres$or_44_pop_use`). Two important interaction effects regulate the frequency-dependence of a player's behavior; if the Fourfour appears in 60% of recent professional games, *and* its population performance is ten percentage points higher than that of its rival, the Threefour, the expected probability a player will use it rises from `r mres$pr_44_pop_use_60`% to `r mres$pr_44_pop_use_60_win_10`% (Fig. \ref{fig:pr44}).

<!-- social knowledge is 1.05x more important -->
<!-- varying effect sigma of 4.00 for player intecepts -->
<!-- varying effect sigma of 9.77 for player slopes on pop knowledge -->

Because the coefficient for personal use (which we call $\beta$) and that for population use ($\gamma$) have the same units, comparing their magnitudes tells us the relative predictive value of personal and population knowledge. For an average player, social knowledge is `r mres$social_indy_ratio` times more important than personal experience in predicting the outcome, but this belies the enormous variation from player to player. Partially pooling games of each Black player, we see a varying effects standard deviation of `r mres$player_indy_varef` for the distribution of player-specific $\beta_j$. For population knowledge, the standard deviation across player intercepts is even larger, `r mres$player_social_varef`. Plotting the joint posterior densities of $\beta_j$ and $\gamma_j$ for each player $j$ (Fig. \ref{fig:playerAgeBetaGamma}, left) reveals that some professionals are substantially better predicted by social knowledge than by their own experiences, while other players are almost completely insensitive to population trends (foremost among them, Takemiya Masaki).

<!-- varying effect age trend -->
<!-- average beta is 2.24 -->

@@

# Discussion

@@

<!-- national differences in sensitivities to social and individual info... -->
<!-- 75 chinese players -->

@@

# Appendix

## The GoGoD Database

@@

<!-- n here same as above -->

@@

## Full Regression Model

@@

<!-- warmup and sampling interation counts -->

@@

>>>>

%Notes to self on compiling without BibTeX
%Compile original file as normal with BibTex, using bibunits
%copy contents of bu1.bbl and paste into bibliography section
%comment out \usepackage{bibunits} in preamble
%comment out begin and end bibunit, and putbib
%compile twice with XeLateX

%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode
\documentclass[reqno,12pt]{amsart}
\usepackage[foot]{amsaddr}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[paperwidth=8.5in,paperheight=11in,text={6in,9in},left=1.25in,top=1in,headheight=0.25in,headsep=0.4in,footskip=0.4in]{geometry}

%citation format
\usepackage[authoryear]{natbib}
%use authoryear in brackets for conventional format
% use "super" in brackets for numbers to biblio

% \bibpunct[, ]{}{}{,}{s}{}{,}
\bibpunct[, ]{(}{)}{,}{a}{}{,}
%change s to a, and put open and closed parentheses in first two bracketed arguments for author-year citations

\usepackage{subfigure}
\usepackage{lineno}
\usepackage{threeparttable}

\usepackage{comment} %make block comments

\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\E}{E}

%get rid of red link boxes
\usepackage{hyperref}
%\usepackage{endfloat}
\hypersetup{
    colorlinks,
    linkcolor={red!80!black},
    citecolor={blue!90!black},
    urlcolor={blue!90!black}
}

%flexible hyperrefs
%\usepackage{cleveref}

%gives ability to fix figure locations with H
\usepackage{float}

%allows subsetting of bibliography units
%\usepackage{bibunits}

\synctex=1

\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
  \renewenvironment{#1}%
     {\linenomath\csname old#1\endcsname}%
     {\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

%\usepackage{lmodern}
%\usepackage{unicode-math}
\usepackage{amsmath}
\usepackage{mathspec}
\usepackage{units}
\usepackage{xltxtra}
\usepackage{xunicode}
\defaultfontfeatures{Mapping=tex-text}
%\setmainfont[Ligatures={Common}]{EB Garamond}
%\setmathrm{EB Garamond}
%\setmathfont(Digits,Latin)[Numbers={Lining,Proportional}]{EB Garamond}

\definecolor{linenocolor}{gray}{0.5}
\renewcommand\thelinenumber{\color{linenocolor}\arabic{linenumber}}

\usepackage{fix-cm}

\setcounter{totalnumber}{10}

\newcommand{\mr}{\mathrm}
\usepackage{hyperref}
\usepackage[normalem]{ulem}
\useunder{\uline}{\ul}{}
\makeatletter
\newenvironment{startquote}[2][2em]
  {\setlength{\@tempdima}{#1}%
   \def\chapquote@author{#2}%
   \parshape 1 \@tempdima \dimexpr\textwidth-2\@tempdima\relax%
   \itshape}
  {\par\normalfont\hfill--\ \chapquote@author\hspace*{\@tempdima}\par\bigskip}
\makeatother
%\renewcommand{\baselinestretch}{1.5}

\input{figures/gameDescriptives.tex}
\input{figures/keyModelCalcs.tex}

\begin{document}
%\begin{bibunit}[naturemag]

%{\abstract \small . }

%\renewcommand*\rmdefault{Minion}

\linenumbers
\modulolinenumbers[5]

\maketitle

%ABSTRACT
%\begin{abstract}\normalsize
\noindent\textsc{Abstract.}
Human culture is widely believed to undergo evolution, via mechanisms rooted in the nature of human cognition. A number of theories predict the kinds of human learning strategies, as well as the population dynamics that result from their action. There is little work, however, that quantitatively examines the evidence for these strategies and resulting cultural evolution within human populations. One of the obstacles is the lack of individual-level data with which to link transmission events to larger cultural dynamics. Here, we address this problem with a rich quantitative database from the East Asian board game known as Go. We draw from a large archive of Go games spanning the last six decades of professional play, and find evidence that the evolutionary dynamics of particular cultural variants are driven by a mix of individual and social learning processes. Particular players vary dramatically in their sensitivity to population knowledge, which also varies by age and nationality. The dynamic patterns of opening Go moves are consistent with an ancient, ongoing arms race within the game itself.
%\end{abstract}

\vspace{12pt}

\vspace{24pt}
% \newpage

@@

In what follows, we present an analysis of such a high-resolution ($n=$ \nGames) archive of cultural patterns: records from the East Asian game of Go. Over many decades of professional play, we can see particular strategies and counterstrategies be innovated, copied, decline, and vanish altogether. Our results provide strong evidence that, on the average, these trends are the result of strategic social learning decisions by professional Go players as they adopt and abandon variant openings. This learning model, in aggregate, is the basis for an apparent evolutionary arms race that is largely responsible for the oscillations in Go openings among professional players. In this way, we are able to link individual-level strategic learning to aggregate population dynamics.

\section{Historical Dynamics in the Game of Go}

@@

\begin{figure}[t]
  \begin{center}
    \includegraphics[scale=1.3]{./assets/kifu.pdf}
    \caption{A complete record, with SGF coordinates, for a 1978 game between Japanese professionals Takemiya Masaki and Kudo Norio. Black (Takemiya) begins with a Fourfour at coordinate \textit{pd} (top right), with White (Kudo) responding at \textit{dd} with another Fourfour. White won by resignation after 232 moves. This game is one of the first successful uses of the \textit{dd} response to the Move 1 Fourfour, and \textit{dd} has subsequently become the dominant response among professionals (see Fig. \ref{fig:m2FreqsWins}).}
    \label{fig:kifu}
  \end{center}
\end{figure}

@@

Beyond this basic outline of the game, there are two key rules regulating play. First, stones or groups of stones of one player that become completely surrounded on the grid by stones of the other are removed from the board (this is the only time stones are moved, once placed). Second, to prevent the possibility of the game entering an infinite loop, the \textit{ko} rule restricts play so that the same board state cannot appear twice in one game.

@@

The first few moves of game of Go, which allow the most possibilites, are actually the most predictable. The opening period of the game is often characterized by highly normative, canalized play, where players often follow one of a set of memorized move sequences called \textit{joseki} (or \textit{jungsuk} in Korea), thought to grant neither player a major advantage. Once a particular \textit{joseki} pattern has started, professional players will draw on an encyclopedic knowledge of explored routes in deciding their next move.

Despite the conservatism of opening play in professional games, novel variations appear regularly and, if widely copied, become new \textit{joseki} in the canon. Go historian Peter Shotwell writes of the evolution of \textit{joseki}: "Some are hundreds (or thousands) of years old; others are born, die, and then are reborn again as the style of play changes; still others are yet to be born and patiently wait their turn to be discovered in some flash of ingenuity or desperation." \citep{shotwell2011go}.

\begin{figure}[t]
  \begin{center}
    \includegraphics[scale=1.2]{./assets/m1FreqsWins.pdf}
    \caption{The ebb and flow of the three major first move variants, with proportion of games won. The Fourfour's rise and fall, here broken into four eras, closely tracks with its relative performance (\textit{top}) since 1967 ($r=$ \corFourFourFrqWin{}). Similarly, the Threethree's brief efflorescence and extinction closely match its relative win rate. Because of its initial dominance, all win rates are expressed relative to the Threefour's, and the dotted gray line indicates the location of the 50\% (break-even) point.}
    \label{fig:m1FreqsWins}
  \end{center}
\end{figure}

\begin{figure}[t]
  \begin{center}
    \includegraphics[scale=1.2]{./assets/m2FreqsWins.pdf}
    \caption{Frequency and performance for four major M2 responses to the Fourfour. Performance for each is calculated as the win rate subtracting the average win rate of alternatives. As at M1, a variants's frequency closely tracks its relative performance. To compare with Fig. \ref{fig:m2FreqsWins}, we define four eras (I to IV) corresponding to the Fourfour's rise and fall.}
    \label{fig:m2FreqsWins}
  \end{center}
\end{figure}

@@

After a brief enthusiasm for the Threethree in the 1960s, the Fourfour rose to dominance, peaking in 1977 (appearing in 78\% of games played) and again in 1996 (82\% of games played). The primary goal of this analysis is to understand why this pattern exists, and what processes are driving evolution at M1.

We can trace the rapid diffusion of the Fourfour during this time back to the success of a single innovator, Japanese professional Takemiya Masaki. Beginning his pro career at the age of 14 in the 1960s, Takemiya has used the Fourfour almost exclusively; of his 520 games as Black in the database, only six do not start with the Fourfour. Takemiya's novel use of this variant in 1968 marks the beginning of its spread within the broader Go community, and his personal performance for the next five years (64\%) also exceeded the average Black win rate (61.5\%).

Subsequent to Takemiya's arrival, the success of the Fourfour appears to have prompted an innovation race in White's Move 2 response. Before 1970, White's response to the Fourfour was almost always a Threefour on an adjacent corner. If we imagine the M1 Fourfour is played in the upper right quadrant (SGF coordinate \textit{pd}), nearly 80\% of the time the M2 response in the late 1960s was coordinate \textit{dc} (Fig. \ref{fig:m2FreqsWins}). By 1977, at the Fourfour's first peak, the \textit{dc} response declined to about 30\% of games. In the interim, M2 variants at \textit{dp}, \textit{cq} and \textit{dd} all became established in rapid sequence. As measured by the proportion of games White won, \textit{dp} and \textit{cq} both consistently outperformed \textit{dc} in the 1970s, blunting the Fourfour's effectiveness and anticipating its decline in use after 1977.

\subsection{Strategic Social Learning}

@@

\section{Methods}

@@

\section{Results}

@@

With all predictors centered on means, the intercept estimates that an average Go player will use the Fourfour about pr44Baseline\% of the time during the study period (i.e. logit(.524)=0.10). A player's recent use of a particular opening move strongly predicts their future use; for an otherwise-average player who used the Fourfour about 60\% of their recent games, the probability they will use the Fourfour rises to pr44BUse60\% (OR: or44BUse{}). There is a smaller (OR: 2.77), but unambiguous, positive performance effect from past success with the Fourfour, adding about two percentage points to the probability of use for a player who has won 10\% more games using the Fourfour than his overall win rate. These two personal effects also interact, so that past success enhances the predictive power of past use (or vis versa).

Most important to the question of social learning and cultural evolution, though, is whether there is predictive value in the recent behavior of \textit{other} players, controlling for individual behavior. Indeed, when the Fourfour appears in 60\% of \textit{all} recent games, the expected probability of use by an average player rises from pr44Baseline\% to pr44PopUse60\% (OR: or44PopUse). Two important interaction effects regulate the frequency-dependence of a player's behavior; if the Fourfour appears in 60\% of recent professional games, \textit{and} its population performance is ten percentage points higher than that of its rival, the Threefour, the expected probability a player will use it rises from pr44PopUse60\% to pr44PopUse60Win10\% (Fig. \ref{fig:pr44}).

Because the coefficient for personal use (which we call $\beta$) and that for population use ($\gamma$) have the same units, comparing their magnitudes tells us the relative predictive value of personal and population knowledge. For an average player, social knowledge is socialIndyRatio{} times more important than personal experience in predicting the outcome, but this belies the enormous variation from player to player. Partially pooling games of each Black player, we see a varying effects standard deviation of playerIndyVaref{} for the distribution of player-specific $\beta_j$. For population knowledge, the standard deviation across player intercepts is even larger, playerSocialVaref. Plotting the joint posterior densities of $\beta_j$ and $\gamma_j$ for each player $j$ (Fig. \ref{fig:playerAgeBetaGamma}, left) reveals that some professionals are substantially better predicted by social knowledge than by their own experiences, while other players are almost completely insensitive to population trends (foremost among them, Takemiya Masaki).

@@

\section{Discussion}

@@

@@

\section{Appendix}

\subsection{The GoGoD Database}

@@

@@

\subsection{Full Regression Model}

@@

@@

\end{document}
